---
title: "World Population"
author: "Dwayne Lobo"
date: "2023-12-03"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(caTools)
library(ROCR)
library(caret)
library(tidyverse)
library(corrplot)

```


```{r}
#getwd()
#setwd("C:/Users/dell/Downloads")
data <- read.csv("Demographic.csv",stringsAsFactors = TRUE)
data <- na.omit(data)
columns_to_drop <- c("Notes", "ISO3_code", "ISO2_code", "SDMX_code", "LocTypeID", "ParentID", "VarID", "LocID")
data <- data[, -which(names(data) %in% columns_to_drop)]
head(data)

```


```{r}
summary(data)
```




```{r}
selected_columns <- data[, 5:20]
library(corrplot)
cor_matrix <- cor(selected_columns)

# Plot the correlation matrix using corrplot
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust")
```
Positive correlations 
Total population (Male and Female included) is strongly positively correlated to NatChange 
Number of births are directly proportional to the total population of the world 

Negative Correlation 
The population growth rate is inversely proportional to Time
CBR is inversely proportional to PopDensity
```{r}
selected_columns <- data[, 20:40]
library(corrplot)
cor_matrix <- cor(selected_columns)

# Plot the correlation matrix using corrplot
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust")
```

```{r}
selected_columns <- data[, 40:59]
library(corrplot)
cor_matrix <- cor(selected_columns)

# Plot the correlation matrix using corrplot
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust")
```



```{r}

library(ggplot2)

# Assuming TPopulation1Jan and MedianAgePop are columns in your dataset
ggplot(data, aes(x = TPopulation1Jan, y = MedianAgePop, color = TPopulation1Jan)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_c() +  # You can choose a different color scale
  labs(title = "Scatter Plot of Population vs. Median Age",
       x = "Population",
       y = "Median Age")


```

```{r}
library(ggplot2)

ggplot(data, aes(x = LocTypeName, y = PopDensity, color = factor(LocTypeName))) +
  geom_boxplot(
    outlier.shape = NA,
    notch = TRUE,
    lwd = 0.5
  ) +
  labs(
    x = "Location Type",
    y = "Population Density",
    title = "Population Density by Location Type",
  ) +
  theme_bw() +
  scale_color_brewer(palette = "Dark2")+ ylim(c(0, 200))

```
Overall Trends:
The box plot reveals a wide range of population densities across different location types.
Certain types, like Development Groups and Income Groups, generally have higher population densities than others, such as SDG Regions and Subregions.
Presence of outliers emphasizes the diversity, representing countries or subregions with notably different population density patterns compared to the median for their location type.


```{r}
library(ggplot2)

world_data_filtered <- data %>%
  filter(TPopulation1Jan > 1000000)

# Create a histogram with colored bars
ggplot(world_data_filtered, aes(x = TPopulation1Jan, fill = cut(TPopulation1Jan, breaks = 30))) +
  geom_histogram(bins = 30, color = "black", alpha = 0.7) +
  scale_fill_viridis_d() +  # You can choose a different color scale
  labs(title = "Histogram of Countries with Population Over 1 Million",
       x = "Population",
       y = "Frequency")


```
```{r}
prop.table(table(data$LocTypeName))
```
Overall Trends:
The histogram portrays a broad spectrum of population sizes among countries with populations over 1 million.
Many countries fall within the 1-2 million population range, while a distinct presence of very large populations is evident in the right tail.
The distribution underscores the diversity in population sizes, ranging from smaller countries to those with substantial populations, such as China and India.



```{r}

library(ggplot2)

# Assign a title to the plot
my_title <- "Your Pie Chart Title"

# Pie chart
pie_chart <- ggplot(data, aes(x = "", fill = as.factor(LocTypeName))) +
  geom_bar(width = 1, stat = "count") +
  coord_polar(theta = "y") +
  theme_minimal() +
  labs(title = my_title) +
  theme(axis.text = element_blank(), axis.title = element_blank(), panel.grid = element_blank())

print(pie_chart)



```



```{r}


# Fit a linear regression model
lm_model <- lm(PopGrowthRate ~ NatChangeRT, data = data)

# Create a scatter plot with the regression line using ggplot2
library(ggplot2)
ggplot(data, aes(x = NatChangeRT , y = PopGrowthRate)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Linear Regression: NatChangeRT vs. PopGrowthRate",
       x = "NatChangeRT",
       y = "PopGrowthRate") +
  theme_minimal()

```
Negative Correlation:
The regression line slopes downward, demonstrating a negative correlation between population growth rate and natural change rate.
Implication: As the natural change rate (births minus deaths) increases, the population growth rate decreases.
Migration Influence:
Net migration plays a role in shaping the relationship between natural change rate and population growth rate.
Impact: Net out-migration of young adults from a country can result in a slower population growth rate, even when the natural change rate is positive.


```{r}


# List of variables to exclude from regression
exclude_variables <- c("SortOrder", "LocTypeName", "Location","Variant")

# List to store regression models
formula <- as.formula(paste("PopGrowthRate ~", paste(setdiff(names(data), c("PopGrowthRate", exclude_variables)), collapse = " + ")))

# Fit the multiple linear regression model
lm_model <- lm(formula, data = data)

# Print summary statistics of the regression model
summary(lm_model)

```
Overall Model Significance:
The regression model is highly significant (p-value < 2.2e-16), indicating that the included predictors collectively explain a large proportion of the variance in the response variable, PopGrowthRate.
Coefficients and Significance:
The coefficients for several predictors are significant, as indicated by the p-values.
The intercept (7.046e-01) and Time (-5.558e-05) are both significant, suggesting that there is a temporal component influencing population growth.
TPopulation1Jan, TPopulation1July, PopSexRatio, MedianAgePop, NatChangeRT, DoublingTime, Births1519, CBR, LE65, LE65Male, LE80, LE80Male, LE80Female, Q0040, Q0040Male, Q0040Female, NetMigrations, and CNMR are also significant predictors.
Direction of Relationships:
Time, TPopulation1Jan, PopSexRatio, MedianAgePop, DoublingTime, LE65, LE65Male, LE80, LE80Male, Q0040, Q0040Male, Q0040Female, and CNMR have negative coefficients, indicating a negative influence on population growth.
TPopulation1July, NatChangeRT, Births1519, CBR, LEx, LExMale, LE15, LE15Male, LE15Female, LE80Female, Q1550, Q1550Male, Q1550Female, Q1560, and Q1560Male have positive coefficients, suggesting a positive influence.
Multicollinearity and Singularities:
Some predictors have singularities (NA values), suggesting potential multicollinearity issues. These include TPopulationFemale1July, PopChange, Deaths, CDR, and LBsurvivingAge1.
Model Fit:
The adjusted R-squared value is high (0.9984), indicating an excellent fit of the model to the data.
The residual standard error is small (0.05383), suggest


```{r}
# Fit a multiple linear regression model
lm_model <- lm(PopGrowthRate ~ Time + TPopulation1Jan + TPopulation1July + DoublingTime+  MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65 + LE65Male + LE80 + LE80Male + LE80Female + NetMigrations + CNMR, data = data)

# Print summary statistics of the regression model
summary(lm_model)
```
Significant Regression Model:
The regression model is highly significant.
Variance Explanation:
The model explains a large proportion (99.84%) of the variance in PopGrowthRate.
Key Predictors:
The following factors are identified as key predictors:
Time
Initial population sizes (January and July)
Doubling time
Median age of the population
Natural change rate
Life expectancy at age 65 (LE65)
Net migrations
Crude net migration rate (CNMR)

```{r}

lm_model <- lm(PopGrowthRate ~ TPopulation1Jan + TPopulation1July + MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65+CNMR , data = data)

# Print summary statistics of the regression model
summary(lm_model)

```
High Significance:
The regression model is highly significant.
Variance Explanation:
The model explains a large proportion (99.84%) of the variance in PopGrowthRate.
Key Variables:
The choice of variables in the model suggests that the following factors play significant roles in explaining population growth rates:
Initial population size Age distribution Natural change rate Doubling time Net migration
```{r}
library(ggplot2)
ggplot(data, aes(x = PopGrowthRate, y = MedianAgePop)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: PopGrowthRate vs. MedianAgePop",
       x = "PopGrowthRate",
       y = "MedianAgePop")
```






The population growth rate of the world is 20% between the period 1950 -2000
It is forecasted that the population growth rate will decrease

```{r}
library(ggplot2)
ggplot(data, aes(x = PopGrowthRate, y = CNMR)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: PopGrowthRate vs. CNMR",
       x = "PopGrowthRate",
       y = "CNMR")
```

The change in net migration rate (per 1000 population) was highest between 1950 – 2000 
It is forecasted that till 2100 the CNMR will normalize

```{r}
library(ggplot2)
ggplot(data, aes(x = Time, y = PopGrowthRate)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: PopGrowthRate vs. Time",
       x = "Time",
       y = "PopGrowthRate")
```


```{r}
library(ggplot2)
ggplot(data, aes(x = Time, y = CNMR)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: CNMR vs. Time",
       x = "Time",
       y = "CNMR")
```


```{r}
data$poprate<-ifelse(data$PopGrowthRate>=2,1,0)
data<- data %>% mutate(poprate = case_when(PopGrowthRate>=2 ~ 1, .default=0))
head(data)
#It is always a good practice to see the proportion of observations we have for each case
table(data$poprate)
prop.table(table(data$poprate))
```

```{r}

library(caret)
set.seed(1013, sample.kind = "Rejection")

#split the dataset leaving 80% of observations in the training dataset and 20% in the test dataset:

index <- createDataPartition(data$poprate, p = 0.7, list = FALSE)
train_df <- data[index, ]
test_df <- data[-index, ]


# Now lets split our dataset into train and test:



```


```{r}
library(rpart)
library(rpart.plot)


tree<-rpart(poprate ~ Time + TPopulation1Jan + TPopulation1July + DoublingTime+  MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65 + LE65Male + LE80 + LE80Male + LE80Female + NetMigrations + CNMR,data=train_df, method="class",cp=0.05)

rpart.plot(tree,digits=-2)


tree2<-rpart(poprate ~ TPopulation1Jan + TPopulation1July + MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65+CNMR,data=train_df, method="class",cp=0.05)

rpart.plot(tree2,digits=-2)

#Time + TPopulation1Jan + TPopulation1July + DoublingTime+  MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65 + LE65Male + LE80 + LE80Male + LE80Female + NetMigrations + CNMR
#TPopulation1Jan + TPopulation1July + MedianAgePop + NatChangeRT + PopGrowthRate + DoublingTime + LE65+CNMR
```
Tree1
40% of the total dataset has a population growth rate of less than 2%
60% of the dataset has population growth rate of less than 2%
40% of the dataset has population growth rate of more than 2%

Tree2
40% of the total dataset has doubling time greater than equal to 35 years
When the doubling time is >= 35 years then the population growth rate is less than 2% which comprises of 60% of the data
Otherwise the population growth rate is greater than 2% and that comprises 40% of the data





```{r}
test_df$pred = predict(tree, newdata = test_df, type="class")
test_df$pred2 = predict(tree2, newdata = test_df, type="class")
```


```{r}
library(caret)
cat("1st tree (using earnings):")
confusionMatrix(test_df$pred,as.factor(test_df$poprate))

cat("2nd tree (using six splits):")
confusionMatrix(test_df$pred2,as.factor(test_df$poprate))


acc_tree <- sum(test_df$pred == test_df$poprate) / nrow(test_df)
acc_tree2 <- sum(test_df$pred2 == test_df$poprate) / nrow(test_df)

print(paste("Accuracy for tree:", round(acc_tree * 100, 2), "%"))
print(paste("Accuracy for tree2:", round(acc_tree2 * 100, 2), "%"))
```
The population growth rate model exhibits remarkable accuracy, effectively classifying instances of both positive and negative growth rates.
The model excels in precision, sensitivity, and specificity, showcasing its reliability in predicting and classifying population growth patterns.
The high performance across metrics indicates the model's robustness in making accurate predictions related to population growth.
It is crucial to acknowledge the exceptional performance while recognizing the need for further validation and consideration of potential biases in the dataset.
These results underscore the model's effectiveness as a valuable tool for understanding and predicting population growth dynamics.
